---
title: "ldaOptim Workflow: Systematic LDA Parameter Optimization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ldaOptim-workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

`ldaOptim` provides a systematic workflow for optimizing Latent Dirichlet Allocation (LDA) topic models. The package implements two-stage parameter optimization: first finding the optimal alpha hyperparameter via cross-validation, then identifying the optimal number of topics using four established metrics. This approach follows the methodology of Jacobi et al. (2016) and provides a modern replacement for the deprecated `ldatuning` package.

The workflow has six steps:

1. Preprocess raw text into a `DocumentTermMatrix`
2. Find optimal alpha via cross-validation (`lda_find_alpha`)
3. Find optimal number of topics (`lda_find_topics`)
4. Fit the final model (`lda_run_models`)
5. Inspect topics (`get_top_words`, `get_frex_words`, `get_top_docs`)
6. Export results (`export_lda_results`)

Steps 2 and 3 include optional numeric elbow-detection helpers (`suggest_alpha_elbow`, `suggest_topics_elbow`) to complement the visual plots.

## Required Packages

```{r setup}
library(ldaOptim)
library(tidytext)
library(dplyr)
library(ggplot2)
library(stopwords)
library(quanteda)
library(topicmodels)
```

---

## Step 0: Preprocess Text Data

All `ldaOptim` functions take a **`DocumentTermMatrix` in topicmodels format** as their primary input. This is different from a raw quanteda DFM or a `tm` package DTM — the conversion step at the end of this section is required.

Your input data should have at minimum two columns: a document index (`index`) and a text column (`text`).

```{r preprocess}
# --- Sampling (adjust n as needed) ---
data3 <- data2 |>
  distinct(text, .keep_all = TRUE) |>
  slice_sample(n = 100000)          # adjust sample size for your corpus

# --- Stopwords ---
mystopwords <- c(
  stopwords("en"),
  stopwords(source = "smart"),
  "posted", "quote", "http", "click", "image",
  "things", "thing", "read", "post",
  "thread", "threads"               # add domain-specific stopwords here
)
mystopwords <- unique(tolower(mystopwords))

# --- Tokenize and clean ---
tidy_data <- data3 |>
  unnest_tokens(word, text) |>
  anti_join(data.frame(word = mystopwords), by = "word") |>
  mutate(nchar = nchar(word)) |>
  filter(nchar > 2) |>                        # remove very short words
  filter(!grepl("[0-9]{1}", word)) |          # remove numbers
  filter(!grepl("\\W", word))                 # remove non-letter characters

# --- TF-IDF filtering ---
# Remove words that appear in too many or too few documents
maxndoc <- 0.5    # word must appear in fewer than 50% of documents
minndoc <- 0.0005 # word must appear in more than 0.05% of documents

templength <- length(unique(tidy_data$index))

good_common_words <- tidy_data |>
  count(index, word, sort = TRUE) |>
  group_by(word) |>
  summarize(doc_freq = n() / templength) |>
  filter(doc_freq < maxndoc) |>
  filter(doc_freq > minndoc)

tidy_data_pruned <- tidy_data |>
  inner_join(good_common_words, by = "word")

# --- Inspect top words before proceeding ---
tidy_data_pruned |>
  group_by(word) |>
  summarise(n = n()) |>
  arrange(desc(n)) |>
  mutate(word = reorder(word, n)) |>
  slice_max(n, n = 50) |>
  ggplot(aes(n, word)) +
  geom_col() +
  labs(title = "Top 50 words", y = NULL)
```

> **Stop here** and inspect the top words plot. If domain-specific noise words appear, add them to `mystopwords` and re-run before continuing.

```{r create-dtm}
# --- Step 1: Create a quanteda DFM from tidy tokens ---
tidy_dfm <- tidy_data_pruned |>
  count(index, word) |>
  cast_dfm(index, word, n)

# --- Step 2: Convert to topicmodels-format DocumentTermMatrix ---
# full_data is a DocumentTermMatrix (DTM) — this is the object passed to 
# all ldaOptim functions (lda_find_alpha, lda_find_topics, lda_run_models)
full_data <- convert(tidy_dfm, to = "topicmodels")

# Verify the object
class(full_data)  # "DocumentTermMatrix"
dim(full_data)    # rows = documents, columns = terms
```

> **Already have a DTM?** If you have an existing `DocumentTermMatrix` from the `tm` package or a previously saved topicmodels object, you can use it directly — skip the preprocessing above and pass your DTM straight to `lda_find_alpha()`. Note that quanteda DFM objects must always be converted first with `convert(dfm, to = "topicmodels")`.

---

## Step 1: Find Optimal Alpha

The alpha hyperparameter controls document-topic density. Lower alpha values produce sparse, focused distributions; higher values produce more diffuse distributions. `lda_find_alpha()` uses k-fold cross-validation across a range of topic numbers to find the alpha that minimizes perplexity.

```{r find-alpha}
alpha_results <- lda_find_alpha(
  dtm             = full_data,       # DocumentTermMatrix from Step 0
  candidate_alpha = c(10, 5, 2, 1),
  candidate_k     = c(2, seq(5, 100, by = 5), 125, 150, 175, 200),
  folds           = 5,
  seed            = 123
)

head(alpha_results)
```

### Visualize Alpha Results

```{r plot-alpha, fig.width=7, fig.height=5}
# Cross-validation perplexity curves by alpha value
plot_alpha_crossval(alpha_results)

# Smoothed perplexity curves
plot_alpha_smooth(alpha_results)

# Second derivative to identify the elbow point for a given alpha
plot_alpha_second_derivative(alpha_results, alpha_value = "1", vline_at = 50)
```

Use the second derivative plot to identify the "elbow" — the K where adding more topics yields diminishing returns. Select your alpha value based on which curve shows the clearest elbow.

### Optional: Numeric Elbow Suggestion

For users who want a numeric companion to the visual plots, `suggest_alpha_elbow()` applies the Kneedle algorithm (Satopää et al., 2011) to the cross-validation curve. For each alpha divisor, it returns the k at the point of maximum perpendicular distance from the chord connecting the first and last points.

```{r suggest-alpha-elbow}
suggest_alpha_elbow(alpha_results)
```

The function returns one row per alpha divisor with:

- `elbow_k`: suggested k at the curve elbow
- `elbow_perplexity`: perplexity at that k
- `chord_distance`: a sharpness diagnostic — larger values indicate a more clearly defined elbow

An overall recommendation (the alpha with the lowest perplexity at its elbow) is printed via `message()`. Elbow detection is inherently approximate — use this as one input alongside the three visual plots, not as a replacement.

---

## Step 2: Find Optimal Number of Topics

With alpha selected, `lda_find_topics()` fits models across a range of K values and calculates four complementary metrics:

| Metric | Direction | Description |
|---|---|---|
| Griffiths2004 | MAXIMIZE | Log-likelihood based |
| CaoJuan2009 | MINIMIZE | Cosine similarity (lower = more distinct topics) |
| Arun2010 | MINIMIZE | KL divergence based |
| Deveaud2014 | MAXIMIZE | Jensen-Shannon divergence |

The optimal K is where the maximize metrics peak and the minimize metrics valley.

```{r find-topics}
topic_results <- lda_find_topics(
  dtm     = full_data,               # DocumentTermMatrix from Step 0
  topics  = seq(10, 200, by = 10),
  method  = "Gibbs",
  control = list(seed = 123, keep = 50),
  verbose = TRUE
)

head(topic_results)
```

### Visualize Topic Metrics

```{r plot-topics, fig.width=7, fig.height=5}
plot_topics_metrics(topic_results)
```

All four metrics are normalized to [0,1] and plotted together. Select the K where metrics converge.

### Optional: Numeric Elbow Suggestion

`suggest_topics_elbow()` applies the same Kneedle approach to each metric curve individually. The method is direction-agnostic — it works identically for metrics that should be minimized (CaoJuan2009, Arun2010) and maximized (Griffiths2004, Deveaud2014).

```{r suggest-topics-elbow}
suggest_topics_elbow(topic_results)
```

The function returns one row per metric with the suggested `elbow_k`, the metric value at that point, and a `chord_distance` sharpness diagnostic. A consensus message prints the range and median of elbow suggestions across metrics.

Because the four metrics frequently disagree, this output is best read together with the `plot_topics_metrics()` plot: metrics with high `chord_distance` values contribute stronger signals, while metrics with low values may have curves too flat or too noisy for reliable elbow detection.

---

## Step 3: Fit Final Model

With optimal alpha and K identified, fit the final model. `lda_run_models()` accepts multiple K values if you want to compare solutions side by side.

```{r run-models}
# alpha = 1 / (k * alpha_divisor)
# e.g. alpha_divisor = 5, k = 50 gives alpha = 0.004
models <- lda_run_models(
  dtm           = full_data,         # DocumentTermMatrix from Step 0
  k_values      = c(40, 50, 60),     # candidate K values from Step 2
  alpha_divisor = 5,
  method        = "Gibbs",
  control       = list(seed = 123)
)
```

---

## Step 4: Inspect Topics

```{r inspect}
# Top words per topic (highest probability)
top_words <- get_top_words(models[[1]], n = 10)
head(top_words)

# FREX words (frequency-exclusivity weighted — more distinctive)
frex_words <- get_frex_words(models[[1]], n = 10)
head(frex_words)

# Top representative documents per topic
top_docs <- get_top_docs(models[[1]], dtm = full_data, n = 5)
head(top_docs)
```

---

## Step 5: Export Results

`export_lda_results()` writes a comprehensive Excel workbook containing model diagnostics, topic-word distributions, and document-topic assignments.

```{r export}
export_lda_results(
  model    = models[[1]],
  dtm      = full_data,
  filename = "lda_results.xlsx"
)
```

---

## References

Jacobi, C., van Atteveldt, W., & Welbers, K. (2016). Quantitative analysis of large amounts of journalistic texts using topic modelling. *Digital Journalism*, 4(1), 89–106.

Murzintcev, N. (2020). *ldatuning: Tuning of the Latent Dirichlet Allocation Models Parameters*. R package.

Satopaa, V., Albrecht, J., Irwin, D., & Raghavan, B. (2011). Finding a "Kneedle" in a Haystack: Detecting Knee Points in System Behavior. *31st International Conference on Distributed Computing Systems Workshops*, 166–171.
